An Experimental Evaluation of Hybrid Querying on Vectors (EA\&B)

### 2025-05-01 VLDB2026/main.tex
modify
Update main.tex

	Despite BigANN 2023 \cite{bigann2023} evaluates the performance of some ANN algorithms, it presents several notable limitations:
	1) The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. 
	\sout{For instance, attribute filtering algorithms such as NHQ and UNG are not included in the evaluation. }
	2) It \textcolor{blue}{focuses only on single-attribute and dual-attribute query scenarios,} 
	\sout{neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes.}\textcolor{blue}{ \sout{For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range) simultaneously when searching.}}
	3) It does not evaluate algorithm performance under different attribute distributions. 
	\sout{However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.}

@@

	\textit{Multi-Thread Search.}
	As shown in Figure~\ref{fig:exp_1_2_SingleLabel_16thread}, UNG continues to achieve the best performance. 
	%Puck performs well on large-scale datasets, but its performance degrades on small datasets. Puck is tailored for large-scale scenarios, and optimizations such as two-level inverted indexing and hierarchical quantization introduce unnecessary overhead on small datasets. 
	\textcolor{blue}{Puck performs well on large datasets but degrades on smaller ones due to overhead from optimizations like two-level inverted indexing and hierarchical quantization. }
	CAPS maintains a relatively high throughput under multi-threads but exhibits limited adaptability to different datasets. The vector distribution of the dataset affects the construction of its partitioned %index.

	ACORN performs poorly overall, primarily due to the sensitivity of its hyperparameter $\gamma$ to attribute selectivity. Without careful tuning based on dataset characteristics, it is challenging to construct an effective index.

	\textit{\textbf{Multi-Attribute Building and Single-Attribute Query.}}
	In this scenario, the index is constructed using 8 attributes but applies filtering condition on only 1 attribute during the query.

	Faiss offers an ID filtering mechanism, it operates solely during query execution and does not influence index building. \textcolor{blue}{ACORN is   designed for} single-attribute indexing. In database systems (VBASE, PASE, Milvus), 
	%\sout{when the system builds both B+ tree and vector indexes, queries typically utilize only the B+ tree.} 
	\textcolor{red}{attributes do not participate in index construction.}
	Hence, they are excluded from this comparison.

@@

	\textit{\textbf{Single-Attribute Selectivity.}}
	In hybrid queries, varying attribute selectivity (AS) can significantly impact computational cost and query efficiency, as the algorithm must efficiently identify data points matching specific attributes within large datasets. AS is defined as the proportion of data points sharing a given attribute value. For example, AS of 1\% indicates that only 1\% of the dataset meets the given attribute.

	To investigate this effect, we evaluate 4 AS \textcolor{blue}{settings} (1\%, 25\%, 50\%, and 75\%) on the SIFT1M dataset \textcolor{blue}{using single-thread execution. We} compare the query performance of attribute filtering algorithms \textcolor{blue}{across these settings}, as shown in Figure~\ref{fig:exp_5_1_1_SingleLabel_1thread}.

	Experimental results show that UNG achieves the best performance across all AS  \textcolor{blue}{settings}, especially at extremely low AS (1\%). 
	This advantage stems from its pre-filtering strategy. Pre-filtering strategy significantly reduces the ANN search space by narrowing down candidates prior to vector computation. In contrast, VBASE performs the worst at 1\% AS due to its post-filtering strategy. Post-filtering strategy first retrieves a large number of candidates and then applies filtering, incurring substantial computational overhead.

@@

			\includegraphics[width=0.95\textwidth]{figures/exp/exp_5_2_1.pdf}
			\caption{Algorithms optimized for low AS.}
			\label{fig:exp_5_2_1}
		\end{subfigure}

@@

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_2.pdf}
			\caption{Algorithms optimized for high AS.}
			\label{fig:exp_5_2_2}
		\end{subfigure}
		\hspace{1mm} % 增加水平间距
		\begin{subfigure}{0.32\textwidth} % 1/3 宽度
			\centering

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_3.pdf}
			\caption{Algorithms insensitive to AS.}
			\label{fig:exp_5_2_3}
		\end{subfigure}

@@

	\textcolor{blue}{Based on the} query performance trends of different algorithms under varying AS \textcolor{blue}{settings}, we categorize the algorithms into 3 groups:

	(1) \textit{Algorithms optimized for low AS.}  
	This group includes ACORN-1, Faiss+HQI\_Batch, StitchedVamana, UNG, Faiss, Puck, and CAPS, as illustrated in Figure~\ref{fig:exp_5_2_1}.

	ACORN-1 performs well in low AS \textcolor{blue}{settings} because fewer nodes share the same attribute. This increases the average degree of retained nodes and improves graph connectivity. CAPS efficiently skips irrelevant subpartitions, reducing computational overhead. However, as AS increases, it must process more subpartitions and handle a larger dataset. StitchedVamana optimizes local adjacency using independent subgraphs, enabling fast localization at low AS. However, at higher AS \textcolor{blue}{settings}, it requires broader global exploration. UNG, Faiss, and Puck all use pre-filtering strategies to shrink the search space in low AS \textcolor{blue}{settings}, leading to better performance.

	Interestingly, Faiss and Faiss+HQI\_Batch perform worst at the 50\% AS instead of 75\%.
	%, which might seem unexpected. 
	This happens because, when AS decreases from 75\% to 50\%, fewer data points are filtered out. As a result, more clusters must be searched, increasing computational cost.

	\par
	(2) \textit{Algorithms optimized for high AS.}  
	This group includes NHQ, FilteredVamana, VBASE, and PASE, as shown in Figure~\ref{fig:exp_5_2_2}.

	At low AS, NHQ struggles to locate relevant regions efficiently, leading to excessive computations on non-matching nodes. FilteredVamana also performs poorly in these scenarios. Its dynamic pruning strategy makes it difficult to balance vector distance and attribute relevance, resulting in overly restricted search paths.
	VBASE relies on post-filtering, which increases computational overhead at low AS. PASE performs poorly at extremely low AS (e.g., 1\%), with recall dropping below 0.2. Its fixed candidate set may not contain enough valid results after filtering, leading to incomplete top-$k$ results and significantly reducing recall.

	\par
	(3) \textit{Algorithms insensitive to AS.}  
	This group includes Milvus and ACORN-\(\gamma\), as illustrated in Figure~\ref{fig:exp_5_2_3}.

	Milvus is mainly limited by system-level communication overhead, such as API latency. As a result, its query performance remains largely unaffected by AS variations. ACORN-\(\gamma\) maintains stable performance across different AS \textcolor{blue}{settings} by using a higher index construction parameter $\gamma$. This ensures a relatively stable average node degree, even after attribute pruning, minimizing the impact of AS changes on query efficiency.

@@

	%Similar to the evaluation of \textit{Attribute Filtering} algorithms, 
	\textcolor{blue}{Similarly}, we also evaluate 3 key metrics mentioned above in \textit{Range Filtering} algorithms.

@@

	\textit{\textbf{Peak memory usage.}}
	As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets. It stores minimal temporary data during index construction.

@@

	%As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets (Deep, YT-Audio). It stores minimal temporary data (e.g., vectors, centroids, distances) during index construction.

@@

	Due to space constraints, Figure~\ref{fig:exp_8_2} reports results for only the largest ($2^{-2}$) and smallest ($2^{-8}$) range ratios. The experimental results show that across all datasets (Deep, YT-Audio, and WIT) and query ranges, WinFilter consistently delivers the best performance. This highlights the robustness of its design, which performs ANN search over only relevant tree nodes and merges results efficiently, making it highly effective for static range query scenarios.

@@

	%	 修改此处
	ACORN performs similarly on both large and small range queries, but its overall performance is worse than the other methods. Unlike other range filtering algorithms, ACORN is not designed with specialized data structures (such as the $\beta$-WST in WinFilter, the rectangle tree and dynamic segment graph in SeRF, or the HSIG in UNIFY) to achieve efficient range filtering. In fact, the graph structure used by ACORN for attribute filtering is the same as that used for range filtering.

	For third-party \textcolor{red}{library} and databases, Figure~\ref{fig:exp_8_2} shows that Faiss performs well in range filtering and even outperforms some specialized range query algorithms in certain cases. Faiss does not include optimizations specifically tailored for range filtering scenarios; its query mechanism is identical to that used in \textcolor{red}{attribute} filtering scenarios. However, since attributes are pre-sorted in range filtering scenarios, the filtering overhead is minimal, leading to improved query performance. Similar to the selectivity observed in attribute filtering, Faiss performs better in small-range filtering scenarios because it searches over the filtered vectors. With a smaller range, fewer points remain after filtering, resulting in lower computational cost. In addition, the performance of Faiss is influenced by data dimensionality; it tends to work better on low-dimensional datasets.

	Vector databases also show relatively poor performance on range queries. Among them, Milvus performs more balanced across both small-range and large-range queries, even surpassing ACORN under high recall settings. PASE performs poorly on small-range queries, with recall lower than 0.1, due to its post-filtering strategy. VBASE performs better on large-range queries. Compared to PASE, its performance on small ranges is significantly improved, achieving higher recall despite very low QPS.

@@

	To evaluate the robustness of the RF-ANNS \textcolor{red}{algorithms} in different scenarios, we \textcolor{red}{add} an OOD dataset Text2Image to the three standard datasets and \textcolor{red}{test} performance under large-range queries. As shown in Figure~\ref{fig:exp_8_3}, most methods \textcolor{red}{perform} worst on the Text2Image dataset. This is because OOD queries make the $k$-nearest neighbor distribution more sparse. As a result, the search needs to visit more nodes, which leads to an exponential growth in the search space and a sharp drop in efficiency.

@@

	However, even though Deep and YT-Audio have similar LID values, the DSG algorithm \textcolor{red}{shows} large performance differences between them. This means that DSG is more sensitive to changes in data distribution. Its robustness is lower than that of other methods, possibly because it depends not only on LID but also on other complex data characteristics.
	%	To examine the robustness of RF-ANNS algorithms, we analyze their performance across different datasets. As shown in Figure~\ref{fig:exp_8_3}, the performance differences. All methods perform worst on WIT, which has the highest LID. On Deep and YT-Audio, which have similar LID, most algorithms achieve comparable performance. 
	%	This suggests that the effectiveness of RF-ANN search is mainly influenced by LID of the dataset. A higher LID inherently increases the difficulty of hybrid query.
	%	
	%	However, the DSG algorithm shows a noticeable performance gap between the Deep and YT-Audio datasets, despite their comparable LID. This indicates that DSG is less robust than other methods when facing dataset-dependent variations, possibly due to its sensitivity to data distribution beyond LID alone.

@@

	Lastly, our findings show that the distribution and cardinality of attributes, as well as dataset properties (e.g., dimensionality and LID), significantly impact the performance of attribute filtering methods. Enhancing the robustness of these algorithms under diverse data conditions is another key challenge that warrants further investigation.

	\section{CONCLUSION}
	\textcolor{blue}{
		In this paper, we evaluate hybrid query methods, including algorithms, vector databases, and libraries. We design experiments to assess the performance of attribute filtering algorithms. We enrich existing datasets with attribute values and create various experimental scenarios to ensure fair comparisons. We run experiments on 7 real-world datasets to analyze the effectiveness of attribute filtering algorithms. We also compare range filtering algorithms using 4 range query datasets. Finally, we analyze the results, summarize key findings, and identify areas for improvement. Our study validates previous research and provides insights for future work.
	}

>>>>

	Despite BigANN 2023 \cite{bigann2023} evaluates the performance of some ANN algorithms, it presents several notable limitations:
	1) The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. 
%	\sout{For instance, attribute filtering algorithms such as NHQ and UNG are not included in the evaluation. }
	2) It \textcolor{blue}{focuses only on single-attribute and dual-attribute query scenarios,} 
%	\sout{neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes.}\textcolor{blue}{ \sout{For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range) simultaneously when searching.}}
	3) It does not evaluate algorithm performance under different attribute distributions. 
%	\sout{However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.}

@@

	\textit{Multi-Thread Search.}
	As shown in Figure~\ref{fig:exp_1_2_SingleLabel_16thread}, UNG continues to achieve the best performance. 
	%Puck performs well on large-scale datasets, but its performance degrades on small datasets. Puck is tailored for large-scale scenarios, and optimizations such as two-level inverted indexing and hierarchical quantization introduce unnecessary overhead on small datasets. 
	\textcolor{blue}{Puck performs well on large datasets but degrades on smaller ones due to overhead from optimizations like two-level inverted indexing and hierarchical quantization. }
	CAPS maintains a relatively high throughput under multi-threads but exhibits limited adaptability to different datasets. The vector distribution of the dataset affects the construction of its \textcolor{blue}{partitioned index.}
	 ACORN performs poorly overall, primarily due to the sensitivity of its hyperparameter $\gamma$ to attribute selectivity. Without careful tuning based on dataset characteristics, it is challenging to construct an effective index.

	\textit{\textbf{Multi-Attribute Building and Single-Attribute Query.}}
	In this scenario, the index is constructed using 8 attributes but applies filtering condition on only 1 attribute during the query process.
	\textcolor{red}{We modified the NHQ source code to support this experiment.}

	Faiss offers an ID filtering mechanism, it operates solely during query execution and does not influence index building. \textcolor{blue}{ACORN is   designed for} single-attribute indexing. In database systems (VBASE, PASE, Milvus), 
	\sout{when the system builds both B+ tree and vector indexes, queries typically utilize only the B+ tree.} 
	\textcolor{red}{The attributes do not participate in index construction.}
	Hence, they are excluded from this comparison.

@@

	\textit{\textbf{Single-Attribute Selectivity.}}
	In hybrid queries, varying levels of attribute selectivity (AS) can significantly impact computational cost and query efficiency, as the algorithm must efficiently identify data points matching specific attributes within large datasets. AS is defined as the proportion of data points sharing a given attribute value. For example, AS of 1\% indicates that only 1\% of the dataset meets the given attribute.

	To investigate this effect, we evaluate 4 AS levels (1\%, 25\%, 50\%, and 75\%) on the SIFT1M dataset \textcolor{blue}{using single-thread execution. We} compare the query performance of attribute filtering algorithms under varying AS conditions, as shown in Figure~\ref{fig:exp_5_1_1_SingleLabel_1thread}.

	Experimental results show that UNG achieves the best performance across all AS levels, especially at extremely low AS (1\%). 
	This advantage stems from its pre-filtering strategy. Pre-filtering strategy significantly reduces the ANN search space by narrowing down candidates prior to vector computation. In contrast, VBASE performs the worst at 1\% AS due to its post-filtering strategy. Post-filtering strategy first retrieves a large number of candidates and then applies filtering, incurring substantial computational overhead.

@@

			\includegraphics[width=0.95\textwidth]{figures/exp/exp_5_2_1.pdf}
			\caption{Algorithms optimized for low AS levels.}
			\label{fig:exp_5_2_1}
		\end{subfigure}

@@

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_2.pdf}
			\caption{Algorithms optimized for high AS levels.}
			\label{fig:exp_5_2_2}
		\end{subfigure}
		\hspace{1mm} % 增加水平间距
		\begin{subfigure}{0.32\textwidth} % 1/3 宽度
			\centering

			\includegraphics[width=0.92\textwidth]{figures/exp/exp_5_2_3.pdf}
			\caption{Algorithms insensitive to AS level.}
			\label{fig:exp_5_2_3}
		\end{subfigure}

@@

	\textcolor{blue}{Based on the} query performance trends of different algorithms under varying AS levels, we categorize the algorithms into 3 groups:

	(1) \textit{Algorithms optimized for low AS levels.}  
	This group includes ACORN-1, Faiss+HQI\_Batch, StitchedVamana, UNG, Faiss, Puck, and CAPS, as illustrated in Figure~\ref{fig:exp_5_2_1}.

	ACORN-1 performs well in low AS scenarios because fewer nodes share the same attribute. This increases the average degree of retained nodes and improves graph connectivity. CAPS efficiently skips irrelevant subpartitions, reducing computational overhead. However, as AS increases, it must process more subpartitions and handle a larger dataset. StitchedVamana optimizes local adjacency using independent subgraphs, enabling fast localization at low AS. However, at higher AS levels, it requires broader global exploration. UNG, Faiss, and Puck all use pre-filtering strategies to shrink the search space in low AS conditions, leading to better performance.

	Interestingly, Faiss and Faiss+HQI\_Batch perform worst at the 50\% AS level instead of 75\%, which might seem unexpected. This happens because, when AS decreases from 75\% to 50\%, fewer data points are filtered out. As a result, more clusters must be searched, increasing computational cost.

	\par
	(2) \textit{Algorithms optimized for high AS levels.}  
	This group includes NHQ, FilteredVamana, VBASE, and PASE, as shown in Figure~\ref{fig:exp_5_2_2}.

	At low AS levels, NHQ struggles to locate relevant regions efficiently, leading to excessive computations on non-matching nodes. FilteredVamana also performs poorly in these scenarios. Its dynamic pruning strategy makes it difficult to balance vector distance and attribute relevance, resulting in overly restricted search paths.
	VBASE relies on post-filtering, which increases computational overhead at low AS. PASE performs poorly at extremely low AS (e.g., 1\%), with recall dropping below 0.2. Its fixed candidate set may not contain enough valid results after filtering, leading to incomplete top-$k$ results and significantly reducing recall.

	\par
	(3) \textit{Algorithms insensitive to AS level.}  
	This group includes Milvus and ACORN-\(\gamma\), as illustrated in Figure~\ref{fig:exp_5_2_3}.

	Milvus is mainly limited by system-level communication overhead, such as API latency. As a result, its query performance remains largely unaffected by AS variations. ACORN-\(\gamma\) maintains stable performance across different AS levels by using a higher index construction parameter $\gamma$. This ensures a relatively stable average node degree, even after attribute pruning, minimizing the impact of AS changes on query efficiency.

@@

	Similar to the evaluation of \textit{Attribute Filtering} algorithms, we also evaluate 3 key metrics mentioned above in \textit{Range Filtering} algorithms.

@@

	\textit{\textbf{Peak memory usage.}}
	As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets. It stores minimal temporary data (e.g., vectors, centroids, distances) during index construction.

@@

	As Figure~\ref{fig:rangeFilter_memory_mb} shows, Faiss uses the least memory on low-dimensional datasets (Deep, YT-Audio). It stores minimal temporary data (e.g., vectors, centroids, distances) during index construction.

@@

	Due to space constraints, Figure~\ref{fig:exp_8_2} reports results for only the largest ($2^{-2}$) and smallest ($2^{-8}$) range ratios. The experimental results show that across all datasets (Deep, YT-Audio, and WIT) and query ranges, WinFilter consistently delivers the best performance.\textcolor{blue}{ This highlights the robustnes} , which performs ANN search over only relevant tree nodes and merges results efficiently, making it highly effective for static range query scenarios.

@@

	%	 修改此处
	ACORN performs similarly on both large and small range queries, but its overall performance is worse than the other methods. Unlike other range filtering algorithms,\textcolor{blue}{ ACORN is not designed with specialized data structures  to achieve efficient range filtering}. In fact, the graph structure used by ACORN for attribute filtering is the same as that used for range filtering.

	For third-party \textcolor{red}{library} and databases, Figure~\ref{fig:exp_8_2} shows that Faiss performs well in range filtering and even outperforms some specialized range query algorithms in certain cases. 
\textcolor{blue}{since attributes are pre-sorted in range filtering scenarios, the filtering overhead is minimal, leading to improved query performance. Similar to the selectivity observed in attribute filtering. }
	Faiss performs better in small-range filtering scenarios because it searches over the filtered vectors. With a smaller range, fewer points remain after filtering, resulting in lower computational cost. In addition, the performance of Faiss is influenced by data dimensionality; \textcolor{blue}{ the higher the dimensionality, the worse the performance.}

	Vector databases also show relatively poor performance on range queries. Among them, Milvus performs more balanced across both small-range and large-range queries, even surpassing ACORN under high recall settings. PASE performs poorly on small-range queries, with recall lower than 0.1, due to its post-filtering strategy. 
	\textcolor{blue}{VBASE performs moderately; it works well for large-range queries, but for small-range queries, although it achieves high recall, its QPS is very low.}ok

@@

	To evaluate the robustness of the RF-ANNS \textcolor{red}{algorithms} in different scenarios, we \textcolor{red}{ add an OOD dataset in the experiments}. As shown in Figure~\ref{fig:exp_8_3}, most methods \textcolor{red}{perform} worst on the Text2Image dataset. This is because OOD queries make the $k$-nearest neighbor distribution more sparse. As a result, the search needs to visit more nodes, which leads to an exponential growth in the search space and a sharp drop in efficiency.

@@

	However, even though Deep and YT-Audio have similar LID values, the DSG algorithm \textcolor{red}{shows} large performance differences between them. This means that DSG is more sensitive to changes in data distribution. 
%	Its robustness is lower than that of other methods, possibly because it depends not only on LID but also on other complex data characteristics.
	%	To examine the robustness of RF-ANNS algorithms, we analyze their performance across different datasets. As shown in Figure~\ref{fig:exp_8_3}, the performance differences. All methods perform worst on WIT, which has the highest LID. On Deep and YT-Audio, which have similar LID, most algorithms achieve comparable performance. 
	%	This suggests that the effectiveness of RF-ANN search is mainly influenced by LID of the dataset. A higher LID inherently increases the difficulty of hybrid query.
	%	
	%	However, the DSG algorithm shows a noticeable performance gap between the Deep and YT-Audio datasets, despite their comparable LID. This indicates that DSG is less robust than other methods when facing dataset-dependent variations, possibly due to its sensitivity to data distribution beyond LID alone.

@@

	Lastly, our findings show that the distribution and the \textcolor{blue}{ number} of attributes, as well as dataset properties (e.g., dimensionality and LID), significantly impact the performance of attribute filtering methods. Enhancing the robustness of these algorithms under diverse data conditions is another key challenge that warrants further investigation.

	\section{CONCLUSION}
	\textcolor{blue}{In this paper, we systematically evaluate hybrid query methods, including various algorithms, vector databases, and libraries. We enrich the dataset with attribute information and design diverse experimental settings. We analyze the performance of these algorithms in depth. We find that these algorithms have different strengths and weaknesses under large-scale data, complex queries, and resource-constrained scenarios. Attribute filtering algorithms show clear limitations in flexibility and Boolean logic support. Range filtering algorithms show weaknesses in multi-attribute queries and high-recall performance. Sometimes, baseline methods like Faiss perform better in high-recall settings. In the future, we need to improve the flexibility and robustness of attribute filtering algorithms and enhance the performance of range filtering algorithms.}
%	\textcolor{blue}{
%		In this paper, we evaluate hybrid query methods, including algorithms, vector databases, and libraries. We design experiments to assess the performance of attribute filtering algorithms. We enrich existing datasets with attribute values and create various experimental scenarios to ensure fair comparisons. We run experiments on 7 real-world datasets to analyze the effectiveness of attribute filtering algorithms. We also compare range filtering algorithms using 4 range query datasets. Finally, we analyze the results, summarize key findings, and identify areas for improvement. Our study validates previous research and provides insights for future work.
%	}
%
